Correctly Optimizing Abstract Abstract Machines

### 2013-03-29 paper/paper.tex
Tightening wording in lazy nondet. section.

\begin{figure}[t]
\begin{center}
\includegraphics[width=3.2in]{church-relative-space}
\includegraphics[width=3.2in]{church-relative-speed}
\includegraphics[width=3.2in]{church-relative-time}
\vspace{-1.5em}
\end{center}
\caption{Factor improvements over the baseline analyzer for the
  \Church{} benchmark in terms of peak memory usage, the rate of state
  transitions, and total analysis time. (Bigger is better.) Each point
  is marked with the section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes interpretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Expressions} & \mexp &=& \svar[^\mlab]\mvar\\
&&|& \slit[^\mlab]\mlit\\
&&|& \slam[^\mlab]\mvar\mexp\\
&&|& \sapp[^\mlab]\mexp\mexp \\
&&|& \sif[^\mlab]\mexp\mexp\mexp \\
\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
\end{array}
\]
\caption{Syntax of ISWIM}
\label{fig:syntax}
\end{figure}

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\mkaddr}\\
&&|& \kar{\mexp,\menv,\mkaddr}\\
&&|& \kif{\mexp,\mexp,\menv,\mkaddr}\\
\text{Addresses} &\maddr&\in&\Addr \\
\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

@@

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
\mstate &\machstep&\mstate' \text{ defined to be the following} \\
&&\text{let } \mcntr' =\tick(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\mkaddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\mkaddr},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \maddr  =\alloc(\mstate) \\
&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

@@

\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\mstate_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\mstate_0} : \wp(\State) \times \wp(\State\times\State) \to \wp(\State) \times \wp(\State\times\State)\text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{align*}
  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
\end{align*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
\end{equation*}
where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.

@@

Since we can cast this optimization as a widening, there is no need to change
the transition relation itself.
%
Rather, what changes is the structure of the fixed-point iteration.
%
In each pass, the algorithm will collect all newly produced stores and join
them together.
%
Then, before each transition, it will install this joined store into current
state.

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \Context \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. For clarity, we will exhibit non-widened
semantics unless a technique specifically requires it. The final step
to the baseline takes the complexity from quartic to cubic in the
monovariant case.

@@

The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

\newcommand{\ext}[3]{#1\sqcup[#2\mapsto#3]}
%\newcommand{\ext}[3]{ext(#1,#2,#3)}

HERE

\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\mkaddr}} \\
\text{ where }
  \maddr_f &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision.

\begin{align*}
\co[^\mcntr]{\kfn{\maddr_f,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr_f)
\end{align*}
Associated with this indirection, we now apply all functions stored in
the address. This nondeterminism is demanded in order to continue with
evaluation. More generally, this is because applied functions are in
\emph{strict position} (more about that later).

@@

The state space changes from one store plus a set of contexts to a set
of seen abstract states (context plus store), $S$, a set of contexts to
step (the frontier), $F$, and the store to step those contexts with, $\msto$.

\begin{equation*}
(S, F, \msto) \mathrel{\widehat{\machstep}} (S \cup S', F', \msto')
\end{equation*}

We constantly see more states, so $S$ is always growing. The frontier,
which is what remains to be done, changes. Let's start with the result
of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):

\begin{equation*}
I = \setof{(c',\msto') \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto'), c \in F}
\end{equation*}

The next store is the least upper bound of all the stores in $I$ (in
order to have only one, and keep the semantics sound):

\begin{equation*}
\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
\end{equation*}

The next frontier is exactly the states that we found from stepping the last frontier, but have not seen before. They must be \emph{states}, so we pair the contexts with the next store:

\begin{equation*}
F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin S}
\end{equation*}

Finally, we add what we know we had not yet seen to the seen set:

\begin{equation*}
  S' = \setof{(c,\msto') \mid c \in F'}
\end{equation*}

To inject a program $e$ into this machine, we start off knowing we
have seen the first state, and that we need to process the first
state:
\begin{align*}
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

\begin{align*}
\Sigma \in \Store^* \qquad S \subseteq {\mathbb N} \times \Context \qquad F \subseteq \Context
\end{align*}
\begin{align*}
(S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
\mbox{where } I &= \setof{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
              (t',\Sigma') &=\left\{\begin{array}{ll}
                           (t+1,\msto'\Sigma') & \text{ if } \msto' \neq \msto \\
                           (t,\Sigma)   & \text{ otherwise}
                          \end{array}\right. \\
              F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
              S' &= \setof{(c,t') \mid c \in F'}\\
\inject(e) &= (\setof{\ttuple{c_0}{0}},\setof{c_0},\bot,\cons{\bot}{\epsilon},0) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

We lift $\dmachstep$ to accommodate for the asymmetry
in the input and output, and change the frontier-based semantics in the following way:

\begin{align*}
(S, F, \msto,\Sigma,t) &\mathrel{\damachstep} (S \cup S', F', \msto',\Sigma',t') \\
\mbox{ where }
 I &= \setof{(c',\msdiff) \mid \ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff} } \\
 \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 \ttuple{t'}{\Sigma'} &=
     \left\{
       \begin{array}{ll}
         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
         \ttuple{t}{\Sigma} & \text{ otherwise}
       \end{array}\right. \\
 F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
 S' &= \setof{(c,t') \mid c \in F'} \\
\appendall(\varnothing) &= \epsilon \\
\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
\end{align*}

@@

\paragraph{Option 2: Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by duplicating the reduction relation to specialize
variable lookup. This works since in the semantics of ISWIM with
store-allocated results consumes the value component of states in one
step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual and implementation complexity for negligible benefit, we
decided against this approach.

>>>>

\begin{figure}[t]
\small
\begin{center}
\includegraphics[width=3.2in]{church-relative-space}
\includegraphics[width=3.2in]{church-relative-speed}
\includegraphics[width=3.2in]{church-relative-time}
\vspace{-1.5em}
\end{center}
\caption{Factor improvements over the baseline analyzer for the
  \Church{} benchmark in terms of peak memory usage, the rate of state
  transitions, and total analysis time. (Bigger is better.) Each point
  is marked with the section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

@@

\begin{figure}[t]
\small
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes interpretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

\begin{figure}
\small
\[
\begin{array}{l@{\qquad}rcl}
\text{Expressions} & \mexp &=& \svar[^\mlab]\mvar\\
&&|& \slit[^\mlab]\mlit\\
&&|& \slam[^\mlab]\mvar\mexp\\
&&|& \sapp[^\mlab]\mexp\mexp \\
&&|& \sif[^\mlab]\mexp\mexp\mexp \\
\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
\end{array}
\]
\caption{Syntax of ISWIM}
\label{fig:syntax}
\end{figure}

\begin{figure}
\small
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\mkaddr}\\
&&|& \kar{\mexp,\menv,\mkaddr}\\
&&|& \kif{\mexp,\mexp,\menv,\mkaddr}\\
\text{Addresses} &\maddr,\mkaddr&\in&\Addr \\
\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

@@

\begin{figure}
\small
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
\mstate &\machstep&\mstate' \text{ defined to be the following} \\
&&\text{let } \mcntr' =\tick(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\mkaddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\mkaddr},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \maddr  =\alloc(\mstate) \\
&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

@@

{\small
\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\mstate_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\mstate_0} : \wp(\State) \times \wp(\State\times\State) \to \wp(\State) \times \wp(\State\times\State)\text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{align*}
  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
\end{align*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
% \begin{equation*}
 $\eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}$
%\end{equation*}
where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.

@@

Since we can cast this optimization as a widening, there is no need to change
the transition relation itself.
%
Rather, what changes is the structure of the fixed-point iteration.
%
In each pass, the algorithm will collect all newly produced stores and join
them together.
%
Then, before each transition, it installs this joined store into current
state.

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
{\small
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \Context \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. For clarity, we will exhibit non-widened
semantics unless a technique specifically requires it. The final step
to the baseline takes the complexity from quartic to cubic in the
monovariant case.

@@

The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.
\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
\text{ where }
  \maddr &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision:
\begin{align*}
\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr)
\end{align*}
Associated with this indirection, we now apply all functions stored in
the address. This nondeterminism is demanded in order to continue with
evaluation. More generally, this is because applied functions are in
\emph{strict position} (more about that later).

@@

The state space changes from a store and set of contexts to a set
of seen abstract states (context plus store), $S$, a set of contexts to
step (the frontier), $F$, and a store to step those contexts with, $\msto$:
\begin{equation*}
(S, F, \msto) \mathrel{\widehat{\machstep}} (S \cup S', F', \msto')
\end{equation*}

We constantly see more states, so $S$ is always growing. The frontier,
which is what remains to be done, changes. Let's start with the result
of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):
%
\begin{equation*}
I = \setof{(c',\msto') \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto'), c \in F}
\end{equation*}
%
The next store is the least upper bound of all the stores in $I$ (in
order to have only one, and keep the semantics sound):
%
\begin{equation*}
\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
\end{equation*}
%
The next frontier is exactly the states that we found from stepping
the last frontier, but have not seen before. They must be
\emph{states}, so we pair the contexts with the next store:
%
\begin{equation*}
F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin S}
\end{equation*}
%
Finally, we add what we know we had not yet seen to the seen set:
\begin{equation*}
  S' = \setof{(c,\msto') \mid c \in F'}
\end{equation*}
%
To inject a program $e$ into this machine, we start off knowing we
have seen the first state, and that we need to process the first
state:
\begin{align*}
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

{\small
\begin{align*}
\Sigma \in \Store^* \qquad S \subseteq {\mathbb N} \times \Context \qquad F \subseteq \Context
\end{align*}
\begin{align*}
(S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
\mbox{where } I &= \setof{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
              (t',\Sigma') &=\left\{\begin{array}{ll}
                           (t+1,\msto'\Sigma') & \text{ if } \msto' \neq \msto \\
                           (t,\Sigma)   & \text{ otherwise}
                          \end{array}\right. \\
              F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
              S' &= \setof{(c,t') \mid c \in F'}\\
\inject(e) &= (\setof{\ttuple{c_0}{0}},\setof{c_0},\bot,\cons{\bot}{\epsilon},0) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}}

@@

We lift $\dmachstep$ to accommodate for the asymmetry
in the input and output, and change the frontier-based semantics in
the following way:
{\small
\begin{align*}
(S, F, \msto,\Sigma,t) &\mathrel{\damachstep} (S \cup S', F', \msto',\Sigma',t') \\
\mbox{ where }
 I &= \setof{(c',\msdiff) \mid \ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff} } \\
 \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 \ttuple{t'}{\Sigma'} &=
     \left\{
       \begin{array}{ll}
         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
         \ttuple{t}{\Sigma} & \text{ otherwise}
       \end{array}\right. \\
 F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
 S' &= \setof{(c,t') \mid c \in F'} \\
\appendall(\varnothing) &= \epsilon \\
\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
\end{align*}}

@@

\paragraph{Option 2: Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by storing the result of the lookup in a value
representing a delayed nondeterministic choice. This has problems with
the next optimization, leading to unsoundness. Regaining soundness
would involve artificial complications to the semantics.

\paragraph{Our choice: option 1}
The configurations that lead to precision loss happen too rarely to
warrant the significant increase in time and memory needed for this
eager non-determinism. Without a widened store, lazy
non-determinism is not a complete abstraction with either approach,
but the precision gains are not immediately apparent due to fake
rebinding (a second reference to the same variable chooses a different
value from the first) and other problems. The performance and memory
gains are apparent, so we favor the simplicity over invisible
gains.}{}
